Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-03-25 endofquarterpaper.qmd
stash

```{r fig.height= 9, fig.width=10}
```

@@

The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.

@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.

@@

@@

These results are supported by the existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. Additionally @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker connections in the Evolved Aurignacian.

It is also possible that the established precedent of aggregation sites could explain some of our results. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites are places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Soffer_Adovasio_2010; @Bourdier_2013; @Svoboda_2013]. It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose. However, without examining environmental data, this remains only a hypothesis at this time.

Our work adds to a growing awareness of diversity within the Upper Paleolithic. Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @3567b0e7-61f1-3f3c-8f91-c78399bc0a5a]. Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @fbdeb6f7-9945-3e64-82d6-7c3ab51f8115; @97387c23-a43b-34e6-95ef-0a248850bcf3; @c35eb420-0df4-3425-adac-d15b922564bd]. This study contributes to that awareness, suggesting permeable, yet distinct cultural patterning and boundaries, as well as patterns of exchange and movement between groups.

@@

Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity. We feel that these patterns do not represent discrete, consistent groups to which all the sites within the group belong, but rather represent metagroups or behavioral types on the ends of a spectrum. This is in accordance with the cutural phylogeny models proposed by @boyd2013cultural, in which cultural groups are defined by presence or absence of certain cultural variants. In this case, restricted sign variability vs. broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.

>>>>

@@

The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.

@@

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose. However, without examining environmental data, this remains only a hypothesis at this time.

@@

These results are supported by the existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. This is supported by other research finding similar evidence of artistic pro

<!-- Add evidence about scholars seeing expansion in signs/artwork following population expansions in other regions/time periods-->

Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

Our work adds to a growing awareness of diversity within the Upper Paleolithic. Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006]. Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @belfer-cohen2003; @bordes1970; @bordes1963]. This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries, as well as patterns of exchange and movement between groups.

This work also adds to our knowledge on connectivity within the Upper Paleolithic. Our network analysis and perMANOVA test results show evolving patterns of connectivity between groups and sites, suggesting dynamic patterns of social exchange and connection.

@@

Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity. We feel that these patterns do not represent discrete, consistent groups to which all the sites within the group belong, but rather represent metagroups or behavioral types on the ends of a spectrum. This is in accordance with the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups. In this case, restricted sign variability vs. broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.
